---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Katie Steele"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
params:
  state: "AZ"
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```


## The question

Every number the American Community Survey publishes is an estimate drawn from a sample, not a count. That means every number comes with a margin of error — a statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.


```{r}
#my_state <- "AZ"        # two-letter abbreviation, in quotes
my_variable <- "B01002_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = params$state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


There are 15 counties in Arizona; the call to `get_acs()` retrieved them all.

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

Greenlee County has the largest relative margin of error at 2.22%. 

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 0.6  ~ "High confidence",
    moe_pct < 1.2 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```


Grouping counties based on `moe_pct` confidence (high: [0-0.6%); moderate: [0.6% - 1.2%); low otherwise), we see that 6 counties have highly reliable estimates, 7 are moderately reliable, and only 2 counties fall into the low-reliability group.


## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row. That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = params$state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 0.6  ~ "High confidence",
    moe_pct < 1.2 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )

```

This table clearly shows that ACS estimates for median age are least reliable in counties with small populations. The relative margin of error is highest in small counties and decreases as counties grow in mean population.


## What this means for policy

We know the least about the two smallest counties in Arizona, Greenlee and La Paz, which make up the "low confidence" `moe_pct` category. The ACS samples 2-3% of U.S. households annually -- and response rates are lower than that -- so in Greenlee, with a population of 9,452, the 5-year estimate is likely based on only a couple hundred households. This leads to a higher sampling error than in larger counties. 

Since sparsely populated counties have the highest sampling error, their estimates are noisier than those in denser areas. Maricopa County, which is one of the largest counties in the United States, has an `moe_pct` of only 0.27%, meaning we're 95% certain that the true median age in Maricopa is within ~1 month of the estimate (37.5 years). Greenlee County has an `moe_pct` of 2.22%; with an estimated median age of 36.2 years, we can only narrow down the true median age to a much wider ~10-month window at the 95% confidence level.

This could have big implications for policymaking. If funding or resource allocation is determined based on a specific cutoff point or threshold, and estimates are treated as equally precise across counties, small counties are most likely to be hurt, since we see more fluctuation and uncertainty in their estimates as a result of smaller sample sizes.

---
